Epistemic Semantics in Guarded String Models

### 2021-01-13 tex/pdraft.tex
fix figure 8 and write KA+H section

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
\usepackage{enumitem}
%\usepackage{caption}
\renewcommand{\UrlFont}{\ttfamily\small}

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  \begin{tabular}[t]{lcc}
    \toprule
        \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule  
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\
        it & $d$ & $d$ \\
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
  \end{tabular}
    \caption{Partial categorial grammar lexicon. The first column has a word form.
    the second column a categorial type, and third column a semantic translation
    in a language that extends the Epik logical language with lambda. The complete grammar
    is in the appendix.}
  \label{lexiconfigure}
\end{figure}

@@

The development here is concerned with defining concrete computable possible
worlds models, and applying them in natural language semantics.  
Issues for further investigation are mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms and coalgebra.

@@

\section*{Appendix}
Complete categorial grammar extending Figure \ref{lexiconfigure}.  The online supplement
includes a version of this grammar that can be run using Barker and Shan's
chart parser for categorial grammar \citep{barker2005parser}.

  \begin{tabular}[t]{lcc}
    \toprule
    \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule
    \multicolumn{3}{c}{{\bf Statives with expletive subject}} \\ 
        heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\ \hline
    \multicolumn{3}{c}{{\bf Agents}} \\ 
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\ \hline
    \multicolumn{3}{c}{{\bf Auxiliary verbs and expletive subjects}} \\        
        it & $d$ & {\em{dummy}} \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        it's & $(t \slash(d \backslash_D t)$ & $\lambda Px.Px$ \\
        isn't & $t\slash(d \backslash_D t)$ & $\lambda Px.\neg P x$ \\  
        doesn't & $(e \backslash t)\slash(d \backslash_V t)$ & $\lambda Px.\neg Px$ \\  \hline        
    \multicolumn{3}{c}{{\bf Tensed attitude verbs}} \\        
        knows & $(e \backslash t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline
    \multicolumn{3}{c}{{\bf Base form attitude verbs}} \\        
        know & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.p\vee{\Diamond}_Rp$\\
        believe & $(e \backslash_V t)\slash_M t$ &  $\lambda pR.{\Diamond}_Rp$\\ \hline       
    \multicolumn{3}{c}{{\bf Complementizers}} \\        
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
    \multicolumn{3}{c}{{\bf Complementizers for base form verbs}} \\        
        that  & $(((e \backslash_V t){\slash_M}t) \backslash (e \backslash_V t)) \slash t$
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        whether  & $(((e \backslash_V t){\slash_M} t) \backslash (e \backslash_V t)) \slash t$ 
        & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
        && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ % \bottomrule    
      \end{tabular}
\end{document}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
\usepackage{enumitem}
\usepackage{mathtools}
%\usepackage{caption}
\renewcommand{\UrlFont}{\ttfamily\small}

@@

\begin{figure}[!t]
  \vspace{1em}
  \centering
  \footnotesize
  %% \begin{tabular}[t]{lcc}
  %%   \toprule
  %%       \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule
  %%       Amy & $e$ & $\hat{R}_a$\\
  %%       Bob & $e$ & $\hat{R}_b$\\
  %%       it & $d$ & $d$ \\
  %%       heads & $d \backslash_D t$ & $\lambda x.\bullet^*h$ \\
  %%       tails & $d \backslash_D t$ & $\lambda x.\bullet^*t$ \\
  %%       is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P~x.~P~x$ \\
  %%       knows & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~p\vee{\Diamond}_Rp$\\
  %%       believes & $(e \backslash t)\slash_M t$ &  $\lambda p~R.~{\Diamond}_Rp$\\
  %%       that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
  %%       & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
  %%       whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$
  %%       & $\lambda p\,m\,R.{\neg}(m\,({\neg}p)\,R)$ \\
  %%       && \multicolumn{1}{r}{$+ {\neg}(m\,p\,R)$} \\ \bottomrule
  %% \end{tabular}

  \newcommand{\hdrspc}{0.3em}
    \begin{tabular}[t]{l>{\hspace{-.8em}}c >{\hspace{-.6em}}c}
    \toprule
    \textsc{Item} & \textsc{Type} & \textsc{Semantics}\\ \midrule
    \multicolumn{3}{c}{{\bf Statives with expletive subject}} \\[\hdrspc]
        heads & $d \backslash_D t$ & $\lambda x.\,\bullet^*h$ \\
        tails & $d \backslash_D t$ & $\lambda x.\,\bullet^*t$ \\ \midrule
    \multicolumn{3}{c}{{\bf Agents}} \\
        Amy & $e$ & $\hat{R}_a$\\
        Bob & $e$ & $\hat{R}_b$\\ \midrule
    \multicolumn{3}{c}{{\bf Auxiliary verbs and expletive subjects}} \\[\hdrspc]
        it & $d$ & {\em{dummy}} \\
        is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P\,x.\,P\,x$ \\
        it's & $(t \slash(d \backslash_D t)$ & $\lambda P\,x.\,P\,x$ \\
        isn't & $t\slash(d \backslash_D t)$ & $\lambda P\,x.\,\neg P\,x$ \\
        doesn't & $(e \backslash t)\slash(d \backslash_V t)$ & $\lambda P\,x.\,\neg P\,x$ \\  \midrule
    \multicolumn{3}{c}{{\bf Tensed attitude verbs}} \\[\hdrspc]
        knows & $(e \backslash t)\slash_M t$ &  $\lambda p\,R.\,p + {\Diamond}_Rp$\\
        believes & $(e \backslash t)\slash_M t$ &  $\lambda p\,R.\,{\Diamond}_Rp$\\ \midrule
    \multicolumn{3}{c}{{\bf Base form attitude verbs}} \\[\hdrspc]
        know & $(e \backslash_V t)\slash_M t$ &  $\lambda p\,R.\,p + {\Diamond}_Rp$\\
        believe & $(e \backslash_V t)\slash_M t$ &  $\lambda p\,R.\,{\Diamond}_Rp$\\ \midrule
    \multicolumn{3}{c}{{\bf Complementizers}} \\[\hdrspc]
        that  & $(((e \backslash t){\slash_M}t) \backslash (e \backslash t)) \slash t$
        & $\begin{pmatrix*}[l]
          \lambda p\,m\,R.\\
          \,\,{\neg}(m\,({\neg}p)\,R)
        \end{pmatrix*}$ \\[1.1em]
        whether  & $(((e \backslash t){\slash_M} t) \backslash (e \backslash t)) \slash t$
        & $\begin{pmatrix*}[l]
          \lambda p\,m\,R.\\
                  \,\, {\neg}(m\,({\neg}p)\,R) \\
          \,\,\,\, +\,{\neg}(m\,p\,R)
        \end{pmatrix*}$\\ \midrule
    \multicolumn{3}{c}{{\bf Complementizers for base form verbs}} \\[\hdrspc]
        that  & $(((e \backslash_V t){\slash_M}t) \backslash (e \backslash_V t)) \slash t$
        & $\begin{pmatrix*}[l]
          \lambda p\,m\,R. \\
          \,\, {\neg}(m\,({\neg}p)\,R)
        \end{pmatrix*}$ \\[1.1em]
        whether & $(((e \backslash_V t){\slash_M} t) \backslash (e \backslash_V t)) \slash t$
        & $\begin{pmatrix*}[l]
          \lambda p\,m\,R.\\
                  \,\, {\neg}(m\,({\neg}p)\,R) \\
          \,\,\,\, +\,{\neg}(m\,p\,R)
        \end{pmatrix*}$ \\
        \bottomrule
      \end{tabular}%%
    \caption{Categorial grammar lexicon. The first column has a word
      form. the second column a categorial type, and third column a
      semantic translation in a language that extends the Epik logical
      language with lambda. The online supplement includes a version
      of this grammar that can be run using Barker and Shan's chart
      parser for categorial grammar \citep{barker2005parser}.}
  \label{lexiconfigure}
\end{figure}

@@

Epik is closely related to Kleene Algebra with Hypotheses\cite{},
\textsc{Ka+H}, which permits user-specification of futher equations
beyond the axioms of \textsc{Ka}. In practice, these hypotheses are
used to express domain-specific properties of the axioms, such as the
commutativity of certain uninterpreted actions in program
verification, such as $x := 3; y := 5 \equiv y := 5; x := 3$). We can
model both the state formulae and the event formulae as \textsc{KA+H}
hypotheses. Specifically, the state formula $\varphi$ can be thought
of as a hypothesis $\varphi \equiv 1$, and an event formula $\zeta_a$
for event $a$, is simply the hypothesis $a \equiv
\sum_{\varphi,\varphi' \in \llbracket \zeta_a \rrbracket}
\varphi;a;\varphi'$. Notwithstanding some early results about
\textsc{Kat} with hypotheses of the form $cp=c$~\cite{}, using basic
\textsc{KA} implementations would rely on using the exponential
encoding of \textsc{Kat} into \textsc{Ka}~\cite{}.

The key facet that differentiates Epik from \textsc{Ka+H} is the
modality. Each alternative relation represents a set of
uni-directional rewrites that must be applied for that agent, rather
than the universal, optional rewrites represented by the \textsc{Ka+H}
hypotheses. It may be possible to encode Epik's alternative relations
into \textsc{Ka+H} hypotheses, but the translation would be
non-trivial, and we leave it as future work.

The development here is concerned with defining concrete computable
possible worlds models, and applying them in natural language
semantics. Issues for further investigation are mathematical
characterizations of epistemic KATs, e.g. sound and complete axioms
and coalgebrae.

@@

%% \section*{Appendix}
%% Complete categorial grammar extending Figure \ref{lexiconfigure}.  The online supplement
%% includes a version of this grammar that can be run using Barker and Shan's
%% chart parser for categorial grammar \citep{barker2005parser}.
